
## Exercise 1 - Discrete random variable

The probability distribution function of a discrete variable k is given by the zero-truncated Poisson distribution:
P(k) = ((λ**k) * e**(−λ) ) / (k! * (1 − e**(−λ)) ) 
for k = 1, 2, 3, . . .

1) Write the R functions for the probability density and cumulative distribution functions, using the R naming convention.
• Assuming λ = 1.4,

```{r}
lambda <- 1.4
x <- seq(0,10)
pois_pdf <- dpois(x,lambda) #pdf
pois_cdf <- ppois(x, lambda) #cdf

```


2) Produce two plots showing the pdf and cdf, separately.

```{r}
barplot(pois_pdf , names=x, col="firebrick2", xlab="x", ylab="f(x)", density=50, main = sprintf("Poisson distr. lambda=%.2f",lambda), ylim=c(0,0.415), cex.lab=1.5, cex.axis=1.25, cex.main=1.25, cex.sub=1.5)
#plot(x,pois_pdf,type="l")
print(pois_pdf)
```

```{r}
barplot(pois_cdf , names=x, col="firebrick2", xlab="x", ylab="f(x)", density=50, main = sprintf("Poisson cdf. lambda=%.2f",lambda), ylim=c(0,1.1), cex.lab=1.5, cex.axis=1.25, cex.main=1.25, cex.sub=1.5)
#plot(x,pois_cdf,type="l")
```


3) Compute the mean value and variance of the probability distribution using R.

```{r}
mean = sum(pois_pdf*x)
std = sum(pois_pdf*(x**2)) - mean**2
cat("mean: ", mean, "\nstd: ", std, "\n")
```



4) Generate a sample of random numbers from this distribution and show them in an histogram.
Evaluate the sample mean

```{r}
sampled_data = rpois(500, lambda=lambda)
breaks = seq(0,10)
hist(sampled_data, col="firebrick2", xlab="x", ylab="frequency", density=50, breaks=breaks, main = sprintf("Distribution of data sampled=%.2f",lambda))

cat("mean sampled data: ", mean(sampled_data),"\nlambda: ", lambda,"\n")


```

## Exercise 1 - Muon's energy distribution 

The energy distribution of CR muons at sea level can be approximated as follows:

p(E) = N*{1 for E < E0, (E − E0 + 1)**(−γ) for E ≥ E0

where E0 = 7.25 GeV and γ = 2.7.

a) Compute the normalisation factor N using R.

```{r}

#RICONTROLLARE####################################################


E0 <- 1
gamma <- 2.7

energy_distr <- function(E, E0, gamma) {

    if (E < E0) {
        return (1)
    }

    else if (E >= E0) {
       return ((E-E0+1)^(-gamma))
    }
}

integral <- integrate(energy_distr, E0 = E0, gamma = gamma, lower = -10 , upper = 10)

N <- 1 / integral

cat("Normalization factor: ", N)


#RICONTROLLARE####################################################


```


b) Plot the probability density function in R.

```{r}

```


c) Plot the cumulative density function in R.

```{r}

```


d) Compute the mean value using R

```{r}

```


e) [Optional] Generate 106 random numbers from this distribution, show them in an histogram and superimpose the pdf (with a line or with a sufficient number of points).
