
## Exercise 1 - Discrete random variable

The probability distribution function of a discrete variable k is given by the zero-truncated Poisson distribution:
P(k) = ((λ**k) * e**(−λ) ) / (k! * (1 − e**(−λ)) ) 
for k = 1, 2, 3, . . .

1) Write the R functions for the probability density and cumulative distribution functions, using the R naming convention.
• Assuming λ = 1.4,

```{r}
library(actuar)

lambda <- 1.4
x <- seq(0,10)
pois_pdf <- dztpois(x,lambda) #pdf
pois_cdf <- pztpois(x, lambda) #cdf

```


2) Produce two plots showing the pdf and cdf, separately.

```{r}
barplot(pois_pdf , names=x, col="firebrick2", xlab="x", ylab="f(x)", density=50, main = sprintf("Poisson distr. lambda=%.2f",lambda), ylim=c(0,0.5), cex.lab=1.5, cex.axis=1.25, cex.main=1.25, cex.sub=1.5)
#plot(x,pois_pdf,type="l")
print(pois_pdf)
```

```{r}
barplot(pois_cdf , names=x, col="firebrick2", xlab="x", ylab="f(x)", density=50, main = sprintf("Poisson cdf. lambda=%.2f",lambda), ylim=c(0,1.1), cex.lab=1.5, cex.axis=1.25, cex.main=1.25, cex.sub=1.5)
#plot(x,pois_cdf,type="l")
```


3) Compute the mean value and variance of the probability distribution using R.

```{r}
mean = sum(pois_pdf*x)
std = sum(pois_pdf*(x**2)) - mean**2
cat("mean: ", mean, "\nstd: ", std, "\n")
```



4) Generate a sample of random numbers from this distribution and show them in an histogram.
Evaluate the sample mean

```{r}
library(actuar)

sampled_data = rztpois(1000, lambda=lambda)
breaks = seq(1,10,1)
hist(sampled_data, col="firebrick2", xlab="x", ylab="frequency", density=50, breaks=breaks, main = sprintf("Distribution of data sampled=%.2f",lambda), xlim=c(0,10))

cat("mean sampled data: ", mean(sampled_data),"\nlambda: ", lambda,"\n")


```





# Exercise 2 - Continuous random variable
• The energy distribution of CR muons at sea level can be approximated as follows
p(E) = N{1 for E < E0; (E − E0 + 1)^(-γ) for E ≥ E0

where E0 = 7.25 GeV and γ = 2.7.

a) Compute the normalisation factor N using R.

```{r}

E0=7.25
gam =2.7

energy_pdf <- function(E) {

    E0=7.25
    gam =2.7

    ifelse(E<E0 , 1, (E - E0 + 1)^(-gam))

}

integral <- integrate(energy_pdf, lower=0 , upper=Inf) #i consider the pdf domain in the lower limit: Energy cannot be 0

N = 1 / integral$value

cat("Normalization factor N = ", N)
```

b) Plot the probability density function in R.

```{r}
energy_pdf_normalized = function(E) {   
    energy_pdf(E) * N
}

x = seq(0,15,0.25)
y = energy_pdf_normalized(x)

plot(x,y, type="l", col="darkblue", xlab="Energy (MeV)", ylab="Probability", ylim=c(0,0.2), main = sprintf("Energy probability distribution"), cex.lab=1.5, cex.axis=1.25, cex.main=1.25, cex.sub=1.5)
```

c) Plot the cumulative density function in R.

```{r}
energy_cdf <- function(x) {
    output = list()
    for (i in x) {
        output = c(output, integrate(energy_pdf_normalized, lower = 0, upper = i)$value)
    }
    return(as.numeric(output))
}

x1 = seq(0,30,0.25)
y1 = energy_cdf(x1)

plot(x1,y1, type="l", col="darkred", xlab="Energy (MeV)", ylab="Probability", ylim=c(0,1), main = sprintf("Energy cdf"), cex.lab=1.5, cex.axis=1.25, cex.main=1.25, cex.sub=1.5)
abline(h = 1, col = "black", lty = 2)  # Adjust color and line type as desired
```

d) Compute the mean value using R

```{r}
f = function(E){
    energy_pdf_normalized(E)*E
}

mean = integrate(f, lower=0, upper = Inf)$value

print(mean)
```

e) [Optional] Generate 1E06 random numbers from this distribution, show them in an histogram and
superimpose the pdf (with a line or with a sufficient number of points)

```{r}
library(GoFKernel)

# Generate random numbers from uniform distribution
n <- 1000  # Number of random numbers to generate
random_numbers <- runif(n)

cdf_inv = inverse(energy_cdf, lower = 0, upper = 1000)

E = sapply(random_numbers, cdf_inv)


hist(E, freq=FALSE, nclass= 19, col="lightgreen", ylim=c(0,0.2), main="Simulated data", xlab="Energy (MeV)")
curve(energy_pdf_normalized, add = TRUE, col="red", lwd=3)
```




# Exercise 3

• Suppose that the average number of accidents at an intersection is two per day.

a) Using Markov’s inequality, find a bound for the probability that at least five accidents will occur
tomorrow.

```{r}
avg = 2 #average number of accidents per day

P_greater5 = avg / 5 #Markov inequality: the probabiìlity of having 5 accidents in a day will be at maximum this value

cat("The upper bound on the probability that at least five accidents will occur
tomorrow is ", P_greater5, "\n")
```

b) Using Poisson random variables, calculate the probability that at least five accidents will occur tomorrow. Compare this value with the bound obtained in the previous point a).

```{r}
poisson_distr = function(n, lambda) {
    return(((lambda^n)*exp(-lambda))/factorial(n))
}

P_1_pois=poisson_distr(1,avg) #avg is equal to lambda since it is the expeted counts in a single time interval (a day)
P_2_pois=poisson_distr(2,avg)
P_3_pois=poisson_distr(3,avg)
P_4_pois=poisson_distr(4,avg)
P_greater5_pois = 1 - P_1_pois - P_2_pois - P_3_pois -P_4_pois 

cat("From poisson distribution: P(n>=5) = ", P_greater5_pois, "\n", "From Markov inequality: P(n>=5) <= ", P_greater5, "\n")
```

The result is consistent with what one could expect: the probability obtained using the poisson distribution respect the Markov's inequality. 
Nevertheless, taking the upper bound of the Markov inequality as an estimate of the true probabilty it seems to be
not a wise choice as the two results are very different from each other.

c) Let the variance of the number of accidents be two per day. Using Chebyshev’s inequality, find a
bound on the probability that tomorrow at least five accidents will occur

```{r}
var = 2

#Chebyshev's inequality
maxP_greater5_cheby = var / (3^2) #the number of events must be at least 3 more than the average of 2 in order to obtain at least 5 events

cat("The upper bound on the probability that at least five accidents will occur
tomorrow is ", maxP_greater5_cheby, "\n")
```

I observe that the result obtained with chebyshev's inequality is much more consistent with the one obtained from 
the poisson distribution

# Exercise 4

The waiting period from the time a book is ordered until it is received is a random variable with mean seven days and standard deviation two days. If Helen wants to be 95% sure that she receives a book by
certain date, how early should she order the book?

*Solution:*

From Chebyshev's inequality

                            P(|x-E[x]| >= k) <= ( var[x] / (k^2) )

with:
 - x the random variable associated to the number of days between the order of the book and its arrival
 - E[x] its expected value, namely  7
 - var[x] its variance, equal to the standard deviation squared, namely 4
 - k is a specific number of days 
this formula give us an upper bound on the probability that the difference between the expected number of days between order and arrive
and the actual number of days waited would be greater than a fixed number (k)

Using the following closure condition:

                        P(|x-E[x]| >= k) + P(|x-E[x]| < k) = 1

We obtain:

                        P(|x-E[x]| < k) = 1 - P(|x-E[x]| >= k) 

And by using the Chebyshev's inequality:

                        P(|x-E[x]| < k) >=  1 - ( var[x] / (k^2) )

Which is what we need in order to solve the problem: finding the number of days k given by the difference between the expected value and 
the actual days of waiting that result in a probability greater than 0.95



```{r}

#initial parameters
mean = 7
std = 2
var = std**2
p = 0.95

#findinng which k ensures a lower bound on probability >0.95
k = seq(1,10,1) #testing days between order and arrive of the book
for (i in k) { 
    lower =  1-(var/i^2) #filling the list with the lower bounds on probability
    if (lower > p) {
        cat("\nFirst k with lower bound greater than", p ,"is k =", i, "\n\n")
        cat("The number of days Helen should order in advance are", i+7+1, "\n\n")
        break  # Stop iteration once the condition is met
    }
}

```

with k=9 we see that we obtain a lower bound of the number of 0.9506173 > 0.95. 
So the number of days before the arrival that Helen should consider in ordering its book are:

                            |x-7| < 9 ----> x < -2 (impossible) or x > 16

In particular, with x=17 Helen will be sure to obtain its book in time with a probability greater 
than 95%




# Exercise 5

An ordinary deck of 52 cards is divided randomly into 26 pairs. Using Chebyshev’s inequality, find an upper bound for the probability that, 
at most, 10 pairs consist of a black and a red card.

*Solution:*

From Chebyshev's inequality

                            P(|x-E[x]| >= k) <= ( var[x] / (k^2) )
with:
 - x the random variable associated to the number of pairs black-red extracted from the 26
 - E[x] the average number of pairs black-red extracted frome the 26
 - var[x] the variance
 - k lower bound on the difference between the pairs extracted vs the average number of pairs extracted

E[x] and var[x] are given by the Hypergeometric distribution

                                             mean = n * B / N
                                             var = (n * B * (N -B) / (N^2))*(1 - (n-1)/(N-1))

with n the number of draws, B the number of black cards in the deck, N the total number of cards

```{r}
#initializing variables
N = 52
B = 26
n = 26


mean = n * B / N
var = (n * B * (N-B) / ( N^2 ) ) * (1 - ( (n-1) / (N-1) ) )

cat("mean =", mean, ",variance =", var, "\n")

upper = var/3^2

cat("Upper bound on probability of obtaining at least 10 red-black pairs out of the total 26:", upper, "\n")

```




# Exercise 6

• In a stationary bus at the departure station, a passenger gets on the bus, on average every 30
seconds.
a) Compute the probability of getting more than 6 passenger after 2 minutes. Evaluate the probability
of having less than 4 passenger after 3 minutes.

```{r}
#using poisson distribution with time dependence
poisson = function(n,delta_t,lambda) { #delta_t is the number of time intervals of 30s you need 
    p = ((lambda*delta_t)^n) * exp(-lambda*delta_t) / factorial(n)
    return(p)
}

lambda = 1/30 #lambda is the average value in a time unit (1s)

p_0 = poisson(0,120,lambda)
p_1 = poisson(1,120,lambda)
p_2 = poisson(2,120,lambda)
p_3 = poisson(3,120,lambda)
p_4 = poisson(4,120,lambda)
p_5 = poisson(5,120,lambda)
p_6plus = 1 - p_0 - p_1 - p_2 - p_3 - p_4 - p_5


p_4minus = poisson(0,180,lambda) + poisson(1,180,lambda) + poisson(2,180,lambda) + poisson(3,180,lambda) 

print(p_6plus)
print(p_4minus)


```


b) Simulate the distribution of the arrival time of the third passenger and superimpose the corresponding pdf.

```{r}
num_sim = 10000
times = rgamma(num_sim, shape = 3, rate = lambda) # the shape parameter represents the number of events (arrivals of passengers)

# Plot the histogram of simulated arrival times
hist(times, freq = FALSE, main = "Arrival times of the third passenger", xlab = "Time (s)")
curve(dgamma(x, shape = 3, rate = lambda), col = "darkblue", add = TRUE) # Exponential distribution

```

there must be something wrong with densities




c) Repeat the procedure of the point b) for the difference in arrival time between the fifth and the
first passenger.

```{r}

```