# ESERCIZIO 1


1) read the data and import in a data.frame or tibble structure
2) merge the five data frames in an unique structure2

```{r}
data1 <- read.csv("/home/albertos/R/Data_CitiBike/JC-201902-citibike-tripdata.csv")
data2 <- read.csv("/home/albertos/R/Data_CitiBike/JC-201903-citibike-tripdata.csv")
data3 <- read.csv("/home/albertos/R/Data_CitiBike/JC-201904-citibike-tripdata.csv")
data4 <- read.csv("/home/albertos/R/Data_CitiBike/JC-201905-citibike-tripdata.csv")
data5 <- read.csv("/home/albertos/R/Data_CitiBike/JC-201906-citibike-tripdata.csv")

data <- rbind(data1,data2,data3,data4,data5)

str(data)
```      


3) check for missing data and remove it, if any

```{r}  

test <- is.na(data)

if (sum(test)>0) {
    print("there are missing data")
} else if (sum(test)==0) {
    print("no missing data")
} else {
    print("error")
}
```      


4.1) compute the average and the median trip duration in minutes

```{r}
mean_td <- mean(data$tripduration) / 60
median_td <- median(data$tripduration) / 60
cat("mean trip duration: ", mean_td, " min\n")
cat("median trip duration: ", median_td, " min\n")
```


4.2) evaluate the minimum and maximum trip duration; does that sound like a reasonable value?

```{r}
max_td = max(data$tripduration) / 60 
min_td = min(data$tripduration) / 60

cat("max trip duration: ", max_td, " min\n")
cat("min trip duration: ", min_td, " min\n")

```

Looks like there is something wrong with max trip duration, it is not reasonable a duration of 28817 minutes (almost 20 days)


4.3) Repeat the calculation of the average (and the median) trip duration by excluding trips longer than
3 hours. Next, evaluate the number of skimmed entries

```{r echo=FALSE}
#loading trip durations
data_skim = data$tripduration

#removing data with trip duration less than 3 hours
map = data_skim < 3*3600 #3600 seconds in an hour
data_skim = data_skim[map]  
cat("Number of skimmed entries: ", sum(map),"\n")

#computing new avg and median
mean_td <- mean(data_skim) / 60
median_td <- median(data_skim) / 60
cat("mean trip duration: ", mean_td, " min\n")
cat("median trip duration: ", median_td, " min\n")

max_td = max(data_skim) / 60 
cat("max trip duration: ", max_td, " min\n")
```


4.4) plot the distribution of trip duration after the skimming of the previous point

```{r}
hist(data_skim/60,  main="Trip durations after skimming",xlab="trip duration (min)",border = "purple", 
     col = "orange",nclass=30)
```

5) plot the monthly average trip duration


```{r}
#creating a list with trip durations for every months (from february 02 to june 06 )
data_td_months = list(data1$tripduration,data2$tripduration,data3$tripduration,data4$tripduration,data5$tripduration)

#filling a list with means trip durations per months
means=list()
for (i in data_td_months) {
    means=c(means, mean(i))
}

#plotting the distribution of average trip durations per month
plot(seq(2,6), as.numeric(means), type="p", col="red", cex=3, pch="+", xlab="months", ylab="Mean trip duration (s)")
```

6.1) plot the number of rides per day

```{r}
if (FALSE) {

    names(data1)

    #grouping data by day
    init=as.Date(format(as.Date(min(data$starttime)), format = "%y-%m-%d 00:00:00 0000"))
    last=as.Date(format(as.Date(min(data$starttime)), format = "%y-%m-%d 23:59:59 9999"))
    one_day<-as.difftime(1, units="days")


    date=init #start with the first recorded day
    rides = list()
    while(date<last) {
        #counting the rides in a specific day
        for (i in data$stoptime) {
            counts=0
            if(as.Date(i) < date+one_day) {
                counts =count+1
            }
            print(counts)
            rides = c(rides,counts) #storing the number of rides in a day

            date = date+one_day #next day
        }
    }

    #plot(seq(1,length(rides), as.numeric(rides), type="p", col="red", cex=3, pch="+")
}
```

```{r}
library(dplyr)

#loading data
data_rides = data 

#Adding a new col with the date (%y-%m-%d) of the stop of the ride
data_rides= mutate(data_rides, day = as.Date(starttime)) 

# Grouping data by the date on which the rides has stopped
data_rides = group_by(data_rides, day) 

#counting the numbers of rides per day
rides = summarise(data_rides, num_rides = n()) #n() is a function that counts the number of observations (rows) in each group

# Print the result
#plot(seq(length(rides$day)), rides$num_rides, xlab= "day", ylab="number of rides", type="p", col="red", cex=3, pch="+")
breaks <- seq(min(data_rides$day), max(data_rides$day) + 4, by="4 days")
hist(data_rides$day,  breaks = breaks,  xlab= "day", ylab="number of rides", cex=3, pch="+", main="rides per day", freq=TRUE, border = "purple", 
     col = "orange")

```


6.2) plot the hourly distribution on weekdays and on weekends

```{r}
library(ggplot2)
library(dplyr)


#loading data
data_rides = data 

# Convert starttime to POSIXct object in order to use function weekdays
data_rides$starttime <- as.POSIXct(data_rides$starttime)

#adding columns for hour and day (litteral) of starttime 
data_rides= mutate(data_rides, day_of_week = weekdays(starttime), hour=format(starttime, format="%H"))

#filtering data distingushing between rides in weekday and weekend
data_rides_weekday = filter(data_rides, day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
data_rides_weekend = filter(data_rides, day_of_week %in% c("Saturday", "Sunday"))

#grouping data by hour of start time
data_rides_weekday_hours = group_by(data_rides_weekday, hour)
data_rides_weekend_hours = group_by(data_rides_weekend, hour)

#counting the number of rides per hour
data_rides_weekday_hours= summarise(data_rides_weekday_hours, num_rides = n())
data_rides_weekend_hours= summarise(data_rides_weekend_hours, num_rides = n())



# Plot the hourly distribution for weekdays
weekday_plot <- ggplot(data_rides_weekday_hours, aes(x = hour, y = num_rides)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Hourly Distribution of Rides on Weekdays",
       x = "Hour of the Day",
       y = "Number of Rides") +
  theme_minimal()

# Plot the hourly distribution for weekdays
weekend_plot <- ggplot(data_rides_weekend_hours, aes(x = hour, y = num_rides)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Hourly Distribution of Rides on Weekend",
       x = "Hour of the Day",
       y = "Number of Rides") +
  theme_minimal()

```

```{r}
# Print the plot
print(weekday_plot)
```
    
```{r}
# Print the plot
print(weekend_plot)
```



6.3) plot again the average hourly distribution on weekdays but separating customer and subscriber
users

CUSTOMERS:
```{r}

library(dplyr)
library(ggplot2)

#filtering weeday data by customers and grouping by hour as before
data_rides_weekday_customer = filter(data_rides_weekday, usertype == "Customer")
data_rides_weekday_customer = group_by(data_rides_weekday_customer, hour)
data_rides_weekday_customer= summarise(data_rides_weekday_customer, num_rides = n())


# Plot the hourly distribution for weekdays for customers
weekday_plot_customer <- ggplot(data_rides_weekday_customer, aes(x = hour, y = num_rides)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Hourly Distribution of Rides on Weekdays for customers",
       x = "Hour of the Day",
       y = "Number of Rides") +
  theme_minimal()

print(weekday_plot_customer)

```

SUBSCRIBERS:
```{r}
library(dplyr)
library(ggplot2)

#filtering weeday data by subscriber and grouping by hour as before
data_rides_weekday_subscriber = filter(data_rides_weekday, usertype == "Subscriber")
data_rides_weekday_subscriber = group_by(data_rides_weekday_subscriber, hour)
data_rides_weekday_subscriber= summarise(data_rides_weekday_subscriber, num_rides = n())


# Plot the hourly distribution for weekdays for subscribers
weekday_plot_subscriber <- ggplot(data_rides_weekday_subscriber, aes(x = hour, y = num_rides)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Hourly Distribution of Rides on Weekdays for subscribers",
       x = "Hour of the Day",
       y = "Number of Rides") +
  theme_minimal()

print(weekday_plot_subscriber)

```



7.1) using the latitude and longitude information3, evaluate the average speed (in km/h) of a user, discarding the trip lasting longer than 1 hour

```{r}
library(geosphere)
library(dplyr)

#loading data
data_distances=data

#discarding trip longer than 1 hour
data_distances = filter(data_distances, tripduration < 3600 ) 

#adding a column with trip distance in m
#the function distHaversine (package geosphere) gives you the shortest distance between two points assuming spherical Earth.
data_distances= mutate(data_distances, tripdistance = distHaversine(p1=cbind(start.station.latitude,start.station.longitude), p2=cbind(end.station.latitude,end.station.longitude)))

#adding a column with avg speed of the trip + conversion factor from m/s to km/h
data_distances= mutate(data_distances, avg_speed = 3.6 * (tripdistance / tripduration ))

print(head(data_distances, 3))
```


7.2) plot the average speed as a function of route length for the following group of distances d < 500 m, 500 m < d < 1000 m, 1000 m < d < 2000 m, 2000 m < d < 3000 m, d > 3000 m and discarding
trips longer than 1 hour

```{r}

library(dplyr)
distances_ranges = c(0,500,1000,2000,3000,max(data_distances$tripdistance))

avg_speeds = list()
for (i in seq(length(distances_ranges)-1) ) {

  #temporary variable
  var = data_distances

  #selecting data with tripdistance in a specific range
  var = filter(var, tripdistance > distances_ranges[i] & tripdistance < distances_ranges[i+1])

  #storing average speed
  avg_speeds[[i]] = mean((var$tripdistance / var$tripduration)*3.6)

} 

barplot(as.numeric(avg_speeds), names = c("[0,500]", "[500,1000]", "[1000,2000]","[2000,3000]", ">3000"), xlab = "Ranges of trip distances (m)", ylab = "Average speed (Km/h)", main = "Avg speed VS Distance")

```



7.3) repeat the same graph, but show the results obtained separately for weekdays and weekends

WEEKDAYS:
```{r}

library(dplyr)
library(ggplot2)
distances_ranges = c(0,500,1000,2000,3000,max(data_distances$tripdistance))

#adding a col with name of the day
data_distances$starttime <- as.POSIXct(data_distances$starttime)
data_distances = mutate(data_distances, day_of_week = weekdays(starttime))

#filtering weekdays
data_distances_weekday = filter(data_distances, day_of_week %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))


avg_speeds = list()
for (i in seq(length(distances_ranges)-1) ) {

  #temporary variable
  var = data_distances_weekday

  #selecting data with tripdistance in a specific range
  var = filter(var, tripdistance > distances_ranges[i] & tripdistance < distances_ranges[i+1])

  #storing average speed
  avg_speeds[[i]] = mean((var$tripdistance / var$tripduration)*3.6)

} 

barplot(as.numeric(avg_speeds), names = c("[0,500]", "[500,1000]", "[1000,2000]","[2000,3000]", ">3000"), xlab = "Ranges of trip distances (m)", ylab = "Average speed (Km/h)", main = "Avg speed VS Distance")

```

WEEKENDS:

```{r}

library(dplyr)
library(ggplot2)
distances_ranges = c(0,500,1000,2000,3000,max(data_distances$tripdistance))

#adding a col with name of the day
data_distances$starttime <- as.POSIXct(data_distances$starttime)
data_distances = mutate(data_distances, day_of_week = weekdays(starttime))

#filtering weekdays
data_distances_weekend = filter(data_distances, day_of_week %in% c("Saturday", "Sunday"))


avg_speeds = list()
for (i in seq(length(distances_ranges)-1) ) {

  #temporary variable
  var = data_distances_weekday

  #selecting data with tripdistance in a specific range
  var = filter(var, tripdistance > distances_ranges[i] & tripdistance < distances_ranges[i+1])

  #storing average speed
  avg_speeds[[i]] = mean((var$tripdistance / var$tripduration)*3.6)

} 

barplot(as.numeric(avg_speeds), names = c("[0,500]", "[500,1000]", "[1000,2000]","[2000,3000]", ">3000"), xlab = "Ranges of trip distances (m)", ylab = "Average speed (Km/h)", main = "Avg speed VS Distance")

```



8.1) find the most common start station and the least popular end station

```{r}
library(dplyr)

start_station = data %>% 
  add_count(start.station.name, sort = TRUE) %>% #sorting number of occurence of starting stations in descending order
  slice_head(n=1) #selecting the most common starting station


end_station = data %>% 
  add_count(end.station.name, sort = TRUE) %>% #sorting number of occurence of ending stations in descending order
  slice_head(n=1) #selecting the most common ending station

print(start_station$start.station.name)
print(end_station$end.station.name)

```

8.2) show the distribution of start stations

```{r}
library(dplyr)
library(ggplot2)

# Counting occurrences of each start station
by_start <- data %>%
  count(start.station.name, sort = TRUE) %>%
  rename(station_name = start.station.name, frequency = n)


# Ploting the bar chart
plot <- ggplot(by_start, aes(x = station_name, y = frequency)) +
  geom_bar(stat = "identity", width = 0.7) +
  labs(x = "Start station name", y = "Frequency") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) #rotating x label to be vertical

print(plot)

```



8.3) find the three most common routes (start and end station) and the three least popular ones

```{r}
library(dplyr)

# Determining the top 3 most common combinations of start and end stations
most_common <- data %>%
  count(start.station.name, end.station.name, sort = TRUE) %>%
  rename(start_station = start.station.name, end_station = end.station.name, frequency = n) %>%
  distinct(start_station, end_station, .keep_all = TRUE) %>%
  arrange(desc(frequency)) %>%
  slice_head(n = 3)

# Determining the bottom 3 least common combinations of start and end stations
least:popular <- data %>%
  count(start.station.name, end.station.name, sort = TRUE) %>%
  rename(start_station = start.station.name, end_station = end.station.name, frequency = n) %>%
  distinct(start_station, end_station, .keep_all = TRUE) %>%
  arrange(desc(frequency)) %>%
  slice_tail(n = 3)


```